---
title: "Species-specific LGTs"
author: "Janina Rinke"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    code_folding: hide
  pdf_document: default
--- 
</STYLE>
---


This script evaluates species-specific LGTs. 
```{r global-options, include=FALSE}
# Define the global chunk options
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

```{r libraries, include = FALSE}
library(knitr)
library(tidyverse)
library(kableExtra)
library(flextable)
library(magrittr)
library(dplyr)
library(ggplot2)
```

### Load the LGT table with all high-quality CDS candidates
```{r}
library(readxl)

# Read CDS level table to extract all unique HGTs later
LGTs<-read_excel("/Users/Janina/sciebo/MASTER/Masterarbeit/lgt/0_data/LGTs.CDSLevel.UniProt.Info.Reannotated.xlsx")
```


```{r}
colnames(LGTs)

# create grouping for protein type
LGTs$type<-"other"
LGTs$type[grep("Ankyrin|OTU|ankyrin|Tetratricopeptide repeat protein|TPR repeat-containing protein|Reverse transcriptase domain-containing protein|Phosphocholine_transferase_AnkX",LGTs$Protein.names)]<-"Ank"
LGTs$type[grep("Cyclopropane|EC 2.1.1.79",LGTs$Protein.names)]<-"CFA"
LGTs$type[grep("N-acetylmuramic",LGTs$Protein.names)]<-"MurNAc"
LGTs$type[grep("DUF1016|Lysozyme|lysozyme|Lyzozyme",LGTs$Protein.names)]<-"Lys"
LGTs$type[grep("Uncharacterized|uncharacterized|unknown|Deleted",LGTs$Protein.names)]<-"Unknown"
LGTs$type[grep("Fragment",LGTs$Protein.names)]<-"Fragmented"
LGTs$type[grep("Transposase|transposase",LGTs$Protein.names)]<-"Transposase"
LGTs$type[grep("50S",LGTs$Protein.names)]<-"Ribosomal protein"

# Remove all non-species specific HGTs
Unique.LGTs<-LGTs %>%
  dplyr::filter(type=="other") %>%                            # filter out all Anks, Lys, MurNAcs, Uncharacterized LGTs...
  dplyr::distinct(Protein.names, .keep_all = T)  %>%          # Only keep unique Proteins
  dplyr::distinct(CDS.UniprotIDs, .keep_all = T) %>%          # Make sure that we only have each UniProtID once
  drop_na(Protein.names)                                      # Remove Proteins with NA

# count number of HGT loci
Unique.loci.count<-Unique.LGTs %>% 
  dplyr::count(locusID)

# count number of species with unique proteins
Unique.species.count<-Unique.LGTs %>% 
  dplyr::count(species, subfamily)

# count the number of bacterial donors
Unique.LGTs$donor<-"other"
Unique.LGTs$donor[grep("Sodalis",Unique.LGTs$Organism)]<-"Sodalis"
Unique.LGTs$donor[grep("Wolbachia",Unique.LGTs$Organism)]<-"Wolbachia"
Unique.LGTs$donor[grep("Serratia",Unique.LGTs$Organism)]<-"Serratia"
Unique.LGTs$donor[grep("Spiroplasma",Unique.LGTs$Organism)]<-"Spiroplasma"
Unique.LGTs$donor[grep("Yersinia",Unique.LGTs$Organism)]<-"Yersinia"
Unique.LGTs$donor[grep("Cardinium",Unique.LGTs$Organism)]<-"Cardinium"
Unique.LGTs$donor[grep("Baumannia|Blochmannia|Candidatus",Unique.LGTs$Organism)]<-"Blochmannia-like"

Unique.bac.count<-Unique.LGTs %>% 
  dplyr::count(donor)
colnames(Unique.bac.count)<-c("donor","count")

head(Unique.LGTs)

# count the number of expressed proteins
Unique.LGTs$Reannotated.CDS.totalcounts<-as.numeric(Unique.LGTs$Reannotated.CDS.totalcounts)

Unique.LGTs %>% 
  dplyr::count(Reannotated.CDS.totalcounts > 100)



#---------------------------------#
#Reannotated.CDS.totalcounts > 100
#FALSE	68			
#TRUE	13			
#NA	7	
#----------------------------------#


# Assign general functions
Unique.LGTs$funct<-"other"
Unique.LGTs$funct[grep("transcriptional|Transcription|transcription",Unique.LGTs$Protein.names)]<-"Transcription"
Unique.LGTs$funct[grep("transposase|Transposase",Unique.LGTs$Protein.names)]<-"TE-related"
Unique.LGTs$funct[grep("transport|Transport",Unique.LGTs$Protein.names)]<-"Transport"
Unique.LGTs$funct[grep("biosynthesis|de novo",Unique.LGTs$Protein.names)]<-"Metabolic function"
Unique.LGTs$funct[grep("biosynthetic|metabolic",Unique.LGTs$Gene.ontology..biological.process.)]<-"Metabolic function"
Unique.LGTs$funct[grep("regulation",Unique.LGTs$Gene.ontology..biological.process.)]<-"Regulatory function"
Unique.LGTs$funct[grep("protein",Unique.LGTs$Gene.ontology..biological.process.)]<-"Protein-related function"
Unique.LGTs$funct[grep("cell wall | membrane",Unique.LGTs$Gene.ontology..biological.process.)]<-"Cell wall/Membrane-related"
Unique.LGTs$funct[grep("cell wall | membrane",Unique.LGTs$Gene.ontology..GO.)]<-"Cell wall/Membrane-related"
Unique.LGTs$funct[grep("DNA|RNA",Unique.LGTs$Gene.ontology..molecular.function.)]<-"DNA repair / DNA binding"
Unique.LGTs$funct[grep("metal",Unique.LGTs$Gene.ontology..molecular.function.)]<-"Metallopeptidase activity"

Funct<-Unique.LGTs %>% 
  dplyr::count(funct)
colnames(Funct)<-c("Function","count")
```

```{r}
# In depth analysis of expressed unique LGTs
Unique.expressed.LGTs<-Unique.LGTs %>%
  dplyr::filter(Reannotated.CDS.totalcounts > 100)

# Plot GO terms for each expressed unique LGT
library(UniprotR)

####### p1
prot_info<-GetProteinGOInfo("A0A060Q018")
p1<-PlotGOAll(GOObj = prot_info, Top = 10, directorypath = getwd(), width = 8, height = 5)
# Customize the title with the protein ID
p1 <- p1 + 
  ggtitle(paste("UDP-N-acetylglucosamine 1-carboxyvinyltransferase"))
p1

####### p2
prot_info<-GetProteinGOInfo("W0HVG4")
p2<-PlotGOAll(GOObj = prot_info, Top = 10, directorypath = getwd(), width = 8, height = 5)
# Customize the title with the protein ID
p2 <- p2 + 
  ggtitle(paste("D-alanine--D-alanine ligase"))
p2

####### p3
prot_info<-GetProteinGOInfo("A0A1A9NKN6")
p3<-PlotGOAll(GOObj = prot_info, Top = 10, directorypath = getwd(), width = 8, height = 5)
# Customize the title with the protein ID
p3 <- p3 + 
  ggtitle(paste("Phosphoribosylformylglycinamidine cyclo-ligase "))
p3

####### p4
prot_info<-GetProteinGOInfo("A0A060Q4G1")
p4<-PlotGOAll(GOObj = prot_info, Top = 10, directorypath = getwd(), width = 8, height = 5)
# Customize the title with the protein ID
p4 <- p4 + 
  ggtitle(paste("DNA mismatch repair protein "))
p4
```
### Make a multi-panel plot
```{r}
library(ggpubr) #load in library for multi-panel figures
library(gridExtra)

p1 <- p1 + theme(plot.margin = unit(c(1, 1, 1, 1), "cm"))
p2 <- p2 + theme(plot.margin = unit(c(1, 1, 1, 1), "cm"))
p3 <- p3 + theme(plot.margin = unit(c(1, 1, 1, 1), "cm"))
p4 <- p4 + theme(plot.margin = unit(c(1, 1, 1, 1), "cm"))

#put all three plots together into one multipanel plot
multi_plot<- ggarrange(p1,p2,p3,p4, #plots that are going to be included in this multipanel figure
                       labels = c("A", "B", "C","D"), #labels given each panel 
                       ncol = 2, nrow = 2, #adjust plot space 
                       common.legend = T #does the plot have a common legend
                       )
#add titles and labels to the multi-panel graph
multi_plot <- annotate_figure(multi_plot,
                              top = text_grob("GO terms", color = "black", face = "bold", size = 11))
multi_plot 

ggsave("/Users/Janina/Documents/MASTER/Masterarbeit/plots/Unique.LGTs.4plots.pdf", multi_plot, width=15, height=10)
```




### Make different plots to provide an overview of species-specific HGTs
```{r}
install.packages("ggpubr")
library(ggpubr)

# Make a plot to show how many species have unique LGTs and they subfamily identity
species_bp <- ggbarplot(Unique.species.count, x = "species", y = "n",
          fill = "subfamily",               # change fill color by cyl
          color = "white",            # Set bar border colors to white
          palette = "jco",            # jco journal color palett. see ?ggpar
          sort.val = "asc",           # Sort the value in ascending order
          sort.by.groups = TRUE,      # Sort inside each group
          x.text.angle = 60           # Rotate vertically x axis texts
          ) +
  ylab("Number of unique HGTs") +
  xlab("")
species_bp + font("x.text", size = 10)

#ggsave("/Users/Janina/Documents/MASTER/Masterarbeit/plots/Thesis_Plots/Unique.LGTs.of.species_RAW.pdf", species_bp, width=15, height=10)

# Make a plot to show the expression 
Unique.LGTs$Reannotated.CDS.lengths<-as.numeric(Unique.LGTs$Reannotated.CDS.lengths)
Unique.LGTs$Reannotated.s_covs<-as.numeric(Unique.LGTs$Reannotated.s_covs)

Unique.LGTs$Uniprot_organism<-gsub("cellular.*?),","",Unique.LGTs$Uniprot_organism)
Unique.LGTs$Uniprot_organism<-gsub(".*?, .*?, .*?, (.*?), .*","\\1",Unique.LGTs$Uniprot_organism)

expression_vs_s_cov <- ggscatter(Unique.LGTs, x = "Reannotated.s_covs", y = "Reannotated.CDS.totalcounts",
                add = "reg.line",               # Add regression line
                conf.int = TRUE              # Add confidence interval
                #color = "Uniprot_organism", palette = "jco", # Color by groups "cyl"
              # shape = "Uniprot_organism"                   # Change point shape by groups "cyl"
                )
expression_vs_s_cov

head(Unique.LGTs)

Unique.LGTs$Reannotated.CDS.totalcounts<-as.numeric(Unique.LGTs$Reannotated.CDS.totalcounts)
# Make a density plot to show s_cov
s_cov<-Unique.LGTs %>%
 #filter( Reannotated.CDS.totalcounts<5000 ) %>%
  ggplot( aes(x=Reannotated.s_covs)) +
    geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.8) +
    theme_minimal()
s_cov
```


## Make a plot for the bacterial donors
```{r}
# Compute percentages
Unique.bac.count$fraction <- Unique.bac.count$count / sum(Unique.bac.count$count)

# Compute the cumulative percentages (top of each rectangle)
Unique.bac.count$ymax <- cumsum(Unique.bac.count$fraction)

# Compute the bottom of each rectangle
Unique.bac.count$ymin <- c(0, head(Unique.bac.count$ymax, n=-1))

# Compute label position
Unique.bac.count$labelPosition <- (Unique.bac.count$ymax + Unique.bac.count$ymin) / 2

# Compute a good label
Unique.bac.count$label <- paste0(Unique.bac.count$donor, "\n n= ", Unique.bac.count$count)

# Make the plot
bac.plot<-ggplot(Unique.bac.count, aes(ymax=ymax, ymin=ymin, xmax=4, xmin=3, fill=donor)) +
  geom_rect() +
  geom_label( x=3.5, aes(y=labelPosition, label=label), size=6) +
  scale_fill_brewer(palette=4) +
  coord_polar(theta="y") +
  xlim(c(2, 4)) +
  theme_void() +
  theme(legend.position = "none")
bac.plot

#ggsave("/Users/Janina/Documents/MASTER/Masterarbeit/plots/Thesis_Plots/Unique.LGTs.bacteria_RAW.pdf", bac.plot, width=10, height=10)
```

# Plot GO terms for Unique LGTs only
```{r}
library(UniprotR)
go<-GetProteinGOInfo(Unique.LGTs$CDS.UniprotIDs[!is.na(Unique.LGTs$CDS.UniprotIDs)])
#Plot Biological process information top 10 go terms  
PlotGOBiological(go, Top = 10) 
#Plot molecular function information top 20 go terms
Plot.GOMolecular(go, Top = 20)
#Plot subcellualr localization information 
Plot.GOSubCellular(go) 
#Combine Gene ontology plots into one plot 
p_combined<-PlotGoInfo(go)
#Handy visualization for publications 
p2_combined<-PlotGOAll(GOObj = go, Top = 10, directorypath = getwd(), width = 8, height = 5)
```


## Make a plot concerning function and identity
```{r}
head(Funct)

Funct$Function<-as.factor(Funct$Function)

Funct %>%
  mutate(Function = fct_reorder(Function, desc(count))) %>%
  ggplot( aes(x=Function, y=count)) +
    geom_bar(stat="identity", fill="#69b3a2", alpha=.6, width=.4) +
    coord_flip() +
    xlab("") +
    theme_minimal()

#ggsave("/Users/Janina/Documents/MASTER/Masterarbeit/plots/Thesis_Plots/Unique.LGTs.function_RAW.pdf", width=12, height=10)
```



## Plot Overview plots for global LGTs
````{r}

LGTs$Reannotated.CDS.lengths<-as.numeric(LGTs$Reannotated.CDS.lengths)
LGTs$Reannotated.s_covs<-as.numeric(LGTs$Reannotated.s_covs)
LGTs$Reannotated.q_covs<-as.numeric(LGTs$Reannotated.q_covs)
LGTs$Reannotated.CDS.totalcounts<-as.numeric(LGTs$Reannotated.CDS.totalcounts)


# Make a density plot to show s_cov
s_cov<-LGTs %>%
  ggplot( aes(x=Reannotated.s_covs)) +
    geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.8) +
    theme_minimal()+
  xlab("Subject coverage of bacterial reference protein")
s_cov

#ggsave("/Users/Janina/Documents/MASTER/Masterarbeit/plots/Thesis_Plots/Scov_global_RAW.pdf", s_cov,width=15, height=10)

# Show q_cov
q_cov<-LGTs %>%
  ggplot( aes(x=Reannotated.q_covs)) +
    geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.8) +
    theme_minimal()
q_cov

#ggsave("/Users/Janina/Documents/MASTER/Masterarbeit/plots/Thesis_Plots/qcov_global_RAW.pdf", q_cov,width=12, height=10)

# Show CDS_lengths (Reannotated)
CDS_lengths<-LGTs %>%
  ggplot( aes(x=Reannotated.CDS.lengths)) +
    geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.8) +
    theme_minimal()
CDS_lengths

#ggsave("/Users/Janina/Documents/MASTER/Masterarbeit/plots/Thesis_Plots/CDS_lengths_global_RAW.pdf", CDS_lengths,width=12, height=10)

expression<-LGTs %>%
  filter(Reannotated.CDS.totalcounts<5000 ) %>%
  ggplot( aes(x=Reannotated.CDS.totalcounts)) +
    geom_density(fill="#69b3a2", color="#e9ecef", alpha=1) +
    theme_minimal()
expression

#ggsave("/Users/Janina/Documents/MASTER/Masterarbeit/plots/Thesis_Plots/expression_global_RAW.pdf", expression,width=12, height=10)

LGTs %>% 
  dplyr::count(Reannotated.CDS.lengths > 6000) # 58 < 6000 bp length

LGTs %>% 
  dplyr::count(Reannotated.CDS.totalcounts > 100) # 176 < 6000 bp length

Expressed.Unique.LGTs<-Unique.LGTs %>% 
  filter(Reannotated.CDS.totalcounts > 100)

Expressed.Unique.LGTs %>% 
  filter(Reannotated.startCodon=="ATG" | Reannotated.startCodon=="TTG" | Reannotated.startCodon=="GTG")
#-------------------------------------------------#
#Reannotated.CDS.totalcounts > 100
#FALSE	557			
#TRUE	384			
#NA	112	
#-------------------------------------------------#

#-------------------------------------------------#
#Reannotated.CDS.totalcounts > 5000
#FALSE	884			
#TRUE	57			
#NA	112
#-------------------------------------------------#
```


### Count number of extended and minimized loci after reannotation
```{r}
LGTs$Reannotated.CDS.lengths<-as.numeric(LGTs$Reannotated.CDS.lengths)
LGTs$CDS.lengths<-as.numeric(LGTs$CDS.lengths)


head(LGTs)
LGTs %>% 
  dplyr::count(Reannotated.CDS.lengths< CDS.lengths)

#--------------------------------------#
#Reannotated.CDS.lengths > CDS.lengths
#FALSE	843			
#TRUE	78			
#NA	132			
#---------------------------------------#

LGTs %>% 
  dplyr::count(CDS.startCodons == "ATG") # without extension 546 ATG start codons

LGTs %>% 
  dplyr::count(Reannotated.startCodon == "ATG") # after extension 628 ATG start codons 

LGTs %>% 
  dplyr::count(CDS.stopCodons == "TAA" | CDS.stopCodons == "TAG") # without extension 586 stop codons
LGTs %>% 
  dplyr::count(Reannotated.stopCodon == "TAA" | Reannotated.stopCodon == "TAG" ) # with extension 704


LGTs %>% 
  dplyr::count(CDS.startCodons == "ATG" & CDS.stopCodons == "TAA" | CDS.stopCodons =="TAG") # 355 complete candidates

LGTs %>% 
  dplyr::count(Reannotated.startCodon == "ATG" & Reannotated.stopCodon == "TAA" | Reannotated.stopCodon =="TAG") #525

````
